### 2014-03-27 reanalysis/ComBat_original/old_rmd/combat_paper_dataset2_reproduced_ttest.rmd
forenkling data set 2

```
Inspection of the covariate/batch balance;
```{r}
print(table(sampleannotation[ , c("Batch", "Type")])  )
```
Three batches and two covariates. There is a "Cell" column in the sample annotation, but it does not seem to have been used.

### Reproduce the original results
Following the description in section A.1 and A.2 and figure texts in the supplementary materials for Johnson et al., we try to reproduce some of their results. First, the heatmap in Figure A.1.
> A heatmap clustering of data set 2. 698 genes with large variation across all the samples are clustered.<cite> Johnson et al.

```{r reproduced_figA1, dev='png', fig.width=8, fig.height=15, fig.cap="fig. 1", tidy=FALSE, results='hold', fig.show='hold'}
variationmeasure = apply(datamatrix, 1, FUN= function(x){var(x)})
clustermatrix = datamatrix[order(variationmeasure, decreasing=TRUE),][1:2698,]
pheatmap(clustermatrix, scale = "row", cluster_rows=T, cluster_cols=T, 
         color = colorRampPalette(c("red", "black", "green"))(n = 299),
         annotation = sampleannotation[, c("Batch", "Type")], 
         cellheight=NA, cellwidth=13,fontsize=12,  drop_levels=TRUE, show_rownames=F,
         main=paste("Reproduced Fig A1.",sep=""),border_color=NA,treeheight_row=0)
rm(variationmeasure, clustermatrix)
```
The heatmap does not look exactly as Fig. A1 in the paper. This could be due to undocumented transformations preformed in Johnson et al., for example log-transformation or standardization. And there are different ways of calculating variation for a probe. What parameters that were used for the clustering could also be different. But the purpose of the figure was to show that a batch effect is present in the data and especially for batch 3. This is also achieved in the reproduced figure.

Next 3 data sets were made,
- **EB2**: Batches 1 and 2 adjusted with non-parametric use of ComBat
- **EB3**: All thee batches adjusted with non-parametric use of ComBat
- **Batch3**: Only batch 3 (no adjustments)
```{r,results='hide', tidy=FALSE}
mat = datamatrix[,sampleannotation$Batch %in% c("1","2")]
EB2 = as.matrix( sva::ComBat(
            dat=mat, 
            batch=sampleannotation[colnames(mat),"Batch"], 
            mod=model.matrix( ~as.factor(sampleannotation[colnames(mat),"Type"])  ), 
            numCovs=NULL, 
            par.prior=useparprior, 
            prior.plots=FALSE))
rm(mat)
EB3 = as.matrix( sva::ComBat(
            dat=datamatrix, 
            batch=sampleannotation[colnames(datamatrix),"Batch"], 
            mod=model.matrix(~as.factor(sampleannotation[colnames(datamatrix),"Type"])), 
            numCovs=NULL, 
            par.prior=useparprior, 
            prior.plots=FALSE))
Batch3 = datamatrix[,sampleannotation$Batch=="3"]

@@

There is no description in how the 770 genes were selected so we use the same approach as in the reproduction of Figure A.1.
```{r reproduced_figA2, dev='png', fig.width=8, fig.height=15, fig.show='hold', tidy=FALSE}
variationmeasure = apply(EB3, 1, FUN= function(x){var(x)})
clustermatrix = EB3[order(variationmeasure, decreasing=TRUE),][1:770,]
pheatmap(clustermatrix, scale = "row", cluster_rows=T, cluster_cols=T, 
         color = colorRampPalette(c("red", "black", "green"))(n = 299),
         annotation = sampleannotation[, c("Batch", "Type", "Cell")],  
         cellheight=NA, cellwidth=13,fontsize=12,  drop_levels=TRUE, show_rownames=F,
         main=paste("Reproduced Fig A2.",sep=""),border_color=NA,treeheight_row=0)
rm(variationmeasure, clustermatrix)
```
Again the heatmap is not exactly as in Johnson et al, but the batch clustering is broken("Batch, bottom annotation row") and the samples cluster more by cell type("Cell", top annotation row) and treatment type("Type", middle annotation row).

>>>>

```
Inspection of the covariate/batch balance;
```{r}
print(table(sampleannotation[ , c("Batch", "Type")])  )
```
Three batches and two covariates. There is a "Cell" column in the sample annotation, but it does not seem to have been used.
<br/>  
<br/>  
### Reproduce the original results
Following the description in section A.1 and A.2 and figure texts in the supplementary materials for Johnson et al., we try to reproduce some of their results. First, the heatmap in Figure A.1.
> A heatmap clustering of data set 2. 698 genes with large variation across all the samples are clustered.<cite> Johnson et al.

```{r reproduced_figA1, dev='png', fig.width=8, fig.height=15, fig.cap="fig. 1", tidy=FALSE, results='hold', fig.show='hold'}
variationmeasure = apply(datamatrix, 1, FUN= function(x){var(x)})
clustermatrix = datamatrix[order(variationmeasure, decreasing=TRUE),][1:2698,]
pheatmap(clustermatrix, scale = "row", cluster_rows=T, cluster_cols=T, 
         color = colorRampPalette(c("red", "black", "green"))(n = 299),
         annotation = sampleannotation[, c("Batch", "Type")], 
         cellheight=NA, cellwidth=13,fontsize=12,  drop_levels=TRUE, show_rownames=F,
         main=paste("Reproduced Fig A1.",sep=""),border_color=NA,treeheight_row=0)
rm(variationmeasure, clustermatrix)
```
The heatmap does not look exactly as [Fig. A.1](http://biostatistics.oxfordjournals.org/content/suppl/2006/04/21/kxj037.DC1/kxj037supp.pdf) in Johnson et al. This could be due to undocumented transformations preformed in Johnson et al., for example log-transformation or standardization. And there are different ways of calculating variation for a probe. The parameters used in the clustering could also be different. But the purpose of the figure was to show that a batch effect is present in the data and especially for batch 3. This is also achieved in the reproduced figure.
<BR/>  
<BR/>  
Next 3 data sets were made,
- **EB2**: Batches 1 and 2 adjusted with ComBat
- **EB3**: All the batches adjusted with ComBat
- **Batch3**: Only batch 3 (no adjustments)
```{r,results='hide', tidy=FALSE}
mat = datamatrix[,sampleannotation$Batch %in% c("1","2")]
EB2 = as.matrix( sva::ComBat(
            dat=mat, 
            batch=sampleannotation[colnames(mat),"Batch"], 
            mod=model.matrix( ~as.factor(sampleannotation[colnames(mat),"Type"])  ), 
            numCovs=NULL, 
            par.prior=useparprior, 
            prior.plots=FALSE))
rm(mat)
EB3 = as.matrix( sva::ComBat(
            dat=datamatrix, 
            batch=sampleannotation[colnames(datamatrix),"Batch"], 
            mod=model.matrix(~as.factor(sampleannotation[colnames(datamatrix),"Type"])), 
            numCovs=NULL, 
            par.prior=useparprior, 
            prior.plots=FALSE))
Batch3 = datamatrix[,sampleannotation$Batch=="3"]

@@

There is no description in how the 770 genes were selected so we use the same approach as in the reproduction of Figure A.1.
```{r reproduced_figA2, dev='png', fig.width=8, fig.height=15, fig.show='hold', tidy=FALSE}
variationmeasure = apply(EB3, 1, FUN= function(x){var(x)})
clustermatrix = EB3[order(variationmeasure, decreasing=TRUE),][1:770,]
pheatmap(clustermatrix, scale = "row", cluster_rows=T, cluster_cols=T, 
         color = colorRampPalette(c("red", "black", "green"))(n = 299),
         annotation = sampleannotation[, c("Batch", "Type", "Cell")],  
         cellheight=NA, cellwidth=13,fontsize=12,  drop_levels=TRUE, show_rownames=F,
         main=paste("Reproduced Fig A2.",sep=""),border_color=NA,treeheight_row=0)
rm(variationmeasure, clustermatrix)
```
Again the heatmap is not exactly as in Johnson et al, but the batch clustering is broken(note, "Batch" is bottom annotation row!) and the samples cluster more by cell type("Cell", top annotation row) and treatment type("Type", middle annotation row).
